Synthesizing Quotient Lenses

### 2018-02-25 papers/QuotientLenses/main.tex
performed suggested surgery

To begin, consider a bidirectional transformation that converts
\bibtex{} citation records such as
\begin{verbatim}
@Book {conway,
Author = "J. H. Conway",
Title = {Regular Algebra and Finite Machines}
}
\end{verbatim}
\noindent
into an equivalent EndNote format like the following
\begin{verbatim}
%0 Book
%T Regular Algebra and Finite Machines
%A J. H. Conway
%F conway
\end{verbatim}
\noindent
and vice versa.
%
Notice that transformations from one format to the other preserve
most of the information present in the data:  the title, the author,
and the label, for instance.  However, some of the syntactic information
may be lost.  For instance, users are typically uninterested in preserving
whitespace between words or
between fields.  The order of author and title fields is also likely
irrelevant and there may be equivalent ways of writing the same
name:  ``Conway, J. H.'' vs ``J. H. Conway.''

%% Moreover,
%% if two \bibtex{} files differ only in ``inessential'' details, we would like
%% to treat them the same by mapping them to identical EndNote representations.
%% In this , the transformation should allow whitespace characters to occur
%% arbitrarily between the words in the label, author, and title fields, as well as
%% in between them. Second, an author's name can have the surname followed by a
%% comma and then the other names separated by arbitrary whitespace, or just all
%% the names separated by arbitrary whitespace (for example, the names ``Conway, J.
%% H.'' and ``J. H. Conway'' are equivalent). Lastly, the transformation should
%% also ignore the order in which the author and title fields occur in each
%% citation record.

When two instances of the same format differ only terms of such inessential
information, it is often useful to treat them as identical and to map them
into canonical representations.
Continuing our example, the following two
\bibtex{} citations represent the same logical object and as a result, should
be translated to the same EndNote record.
\begin{verbatim}
@Book {conway,
Author = "J. H. Conway",
Title = {Regular Algebra and Finite Machines},
}

@Book{     conway,
Title = {Regular Algebra
and Finite Machines},
Author = "Conway, J. H.",     }
\end{verbatim}

@@

%% %Foster's Boomerang~\ref{boomerang} language presents one design
%% While it is possible to express the equivalence on the \bibtex{} data above in
%% Boomerang, actually doing so is a challenge.  Three of the inessential details
%% we wish to ignore---whitespace quotienting
%% inside fields, whitespace quotienting in between fields, and permutation
%% quotenting of fields---nest within each other in a way that makes them difficult
%% to express with the existing Boomerang combinators. Furthermore, if the
%% programmer wishes to specify the type of the lens, then she can only specify
%% data formats on each side of the lens with a regular expression that does not
%% keep track of the equivalence relation defined on the language of the regular
%% expression. Consequently, the programmer has to write a regular expression as
%% well as a canonizer that expresses the equivalence relation on that regular
%% expression, even though a regular expression and the canonizer defined on it
%% typically share much of the syntactic structure. Having to write both of these
%% expressions makes programming with quotient lenses verbose and time consuming.

@@

\begin{lstlisting}
let wsp = project [ \n\t\r]$^+$ $\to$ " "
\end{lstlisting}

Sometimes there are multiple disjoint representations of the same data.
In such situations, the \kw{squash} combinator creates a QRE that
allows external data to be in either format, and converts any
data in one format to the other.
For instance, assume
the \codefont{comma\_name} format describes ``Conway, J. H.''
and the \codefont{space\_name} format describes ``J. H. Conway''
and \codefont{s\_to\_c} is a function from the first to the second.  In this case,
the following instance of \kw{squash} creates the desired canonizer.

@@

One way to define the function used by squash is simply to write it
from scratch in OCaml.  However, our system will also synthesize
such functions automatically from pairs of regular languages and
examples.
%
\begin{lstlisting}
let l = synth comma_name $\Leftrightarrow$ space_name using {("Conway, J. H.", "J. H. Conway")}
let s_to_c = get l
\end{lstlisting}
%
The first line above synthesizes a \emph{bidirectional} lens between
\codefont{comma\_name} and \codefont{space\_name} using the listed example
transformation as a guide.  The second line simply
extracts the get direction transformation from the lens,
which is what we need for squash.

A third useful QRE primitive is the permutation combinator.  For example,
the following instance of \kw{perm} allows label, author and title fields
(which we assume have been defined earlier) to appear in any order.
%
\begin{lstlisting}
let bib_fields = perm (label, bib_author, bib_title)
\end{lstlisting}
%
To normalize any repeated separator between fields, one can extend a permutation
with any other QRE.  For instance, below, we normalize whitespace between fields,
leaving only a single newline.
%
\begin{lstlisting}
let bib_fields = perm (label, bib_author, bib_title) with (project ("," .wsp$^+$) $\to$ ",\n")
\end{lstlisting}
%
\noindent
The QRE for the bibtex format is presented in its entirety in Figure~\ref{fig:example-qre}.

\begin{figure}[t]
\begin{lstlisting}
(* whitespace with " " as a canonical representative *)
let wsp = project [ \n\t\r]$^+$ $\to$ " "

(* define name representations with a space and with a comma *)
let space_name = name | (name . wsp . (name . wsp)$^*$ . name)
let comma_name = name | (name . "," . wsp . (name . wsp)$^*$ . name)

(* synthesize a lens that maps the comma representation to the space representation *)
let l = synth comma_name $\Leftrightarrow$ space_name using {("Conway, J. H.", "J. H. Conway")}
let s_to_c = get l

(* squash QRE maps comma_name to space_name *)
let name = squash comma_name $\to$ space_name using s_to_c

(* define rest of bibtex fields *)
let name_sep = wsp . "and" . wsp
let bib_names = name | (name . name_sep . (name . name_sep)$^*$ . name)
let bib_author = "author = \"" . bib_names . "\""
let title = word | (word . wsp . (word . wsp)$^*$ . word)
let bib_title = "title = {" . title . "}"

(* allow any permutation of fields interspersed with arbitrary whitespace *)
let bib_fields = perm (label, bib_author, bib_title) with (project ("," .wsp$^+$) $\to$ ",\n")
let bibtex = "@book{" . bib_fields . "}"
\end{lstlisting}
\caption{QRE specification of \bibtex{} records. }
\label{fig:example-qre}
\end{figure}

%% In Figure~\ref{fig:example-qre}, the QRE \verb+project [ \n\t\r]+$^{+} \to$
%% \verb+" "+ matches any non-zero number of whitespace characters and quotients
%% them to a single equivalence class with a single space as the canonical
%% representative. We call this QRE a \textit{projection} because an entire regular
%% language is projected to a single representative string.

%% Another QRE in Figure~\ref{fig:example-qre} is given by

%% \begin{center}
%% \begin{lstlisting}
%% squash comma_name $\to$ space_name using (get space_to_comma)
%% \end{lstlisting}
%% \end{center}

%% \noindent This QRE is a \textit{squash} QRE. A \textit{squash} QRE ``glues''
%% two data formats using a user-defined function. For example,
%% the QRE in the code above specifies that the \cd{comma_name} representation of
%% the author's name is equivalent to the \cd{space_name} representation, with
%% \cd{space_name} taken to be the canonical representation. The
%% function that identifies the two formats is given by \cd{get space_to_comma},
%% which transforms names in the \cd{comma_name} representation to names in the
%% \cd{space_name} representation. In this example, \cd{space_to_comma} is itself
%% an automatically \textit{synthesized} lens between the \cd{space_name} and
%% \cd{comma_name} QREs, and we need to use the ``get'' direction of the lens to
%% specify the squash transformation.

%% The third example of a QRE in the Figure is \cd{bib_fields}, defined by:
%% \begin{center}
%% \begin{lstlisting}
%% perm (label, bib_author, bib_title) with (project ("," . wsp$^+$ ) $\to$ ",\n")
%% \end{lstlisting}
%% \end{center}

%% \noindent This is a \textit{permutation} QRE that specifies that the label,
%% title, and author fields can occur in any order, with the fields separated by a
%% comma followed by arbitrary amounts of whitespace. The canonical representation
%% is that in which the label, author and title fields occur in that order,
%% separated by a comma followed by a newline. The \textit{permutation} QRE is
%% designed for equivalences such as this, where different permutations of
%% equivalences are interspersed with a separator which itself has an
%% equivalence defined on it.

%% Given the \textit{projection}, \textit{squash}, and \textit{permutation}
%% QRE primitives, we can build more complex languages using the usual regular
%% expression combinators of alternative choice `\cd{|}', concatenation `\cd{.}'
%% and Kleene star `$^*$', and also sequential composition ``;''.

\subsection{QRE Lenses}
QREs allow users to specify regular languages modulo equivalence relations,
a first step in writing many data transformations.  However, writing the
inner bijective lens itself still takes a fair amount of work.  Each such lens
is constructed from simple primitive combinators including:
%
\begin{itemize}
  \item \kw{del} $s$: delete the constant string $s$ in the get direction;
  produce the constant string $s$ in the put direction.
  \item \kw{ins} $s$: produce the string $s$ in the get direction;
  delete the string $s$ in the put direction.
  \item \kw{copy} $R$: copy the text matching the regular expression $R$ in both
  directions.
\end{itemize}
In addition, such primitives may be combined using the regular operators
Kleene star, alternation and concatenation, or functional composition of
lenses.  Figure~\ref{fig:example-lens} illustrates the use of these
combinators to transform \bibtex{} to EndNote (and vice versa).
A fair amount of trickiness in writing these transformations as users
must keep mental track of what they have parsed so far \emph{in both
directions}, carefully inserting and deleting the relevant syntax on both
sides.

@@

let author_lens : (lens in (",\n" . bib_author) $\Leftrightarrow$ ("\n" . end_author)) =
    del ",\nauthor = \""
        . ins "\n%A "
        . valid_name
        . (del " and "
        . ins "\n"
        . (ins "%A " . valid_name . del " and " . ins "\n")* 
        . ins "%A "
        . valid_name
        || "")

@@

let bib_to_end : (lens in bibtex $\Leftrightarrow$ end_note) =
    preamble . label . author_lens . title_lens
\end{lstlisting}
\caption{An example lens between \cd{bibtex} and \cd{end_note} regular
expressions with no equivalences.\bcp{Indentation needs work.  And there was
a suggestion to colorize the new keywords that we're introducing here
(including symbolic ones, I guess)}}
\label{fig:example-lens}
\end{figure}

%% The lens shown in Figure~\ref{fig:example-lens} transforms \bibtex{} data to
%% EndNote data  (and vice versa) by matching the delimiters and separators in a
%% bijective manner. For example, the \cd{preamble} lens replaces the \bibtex{}
%% preamble \verb|@Book| with the EndNote preamble \verb|%0 Book\n%F|. When the
%% programmer is writing the lens, she has to make sure that all of these small
%% deletions and insertions are correct, which can be a challenging task if the
%% lens is large enough, as in the definition of the lens \verb!author_lens! in
%% Figure~\ref{fig:example-lens}. This makes lens programming tedious and
%% error-prone.

Consequently, we offer the programmer some relief by identifying a class of
quotient lenses that we call {\em QRE lenses}, and by showing how to {\em synthesize}
QRE lenses from a pair of QREs and a set of input-output example pairs. Our key
insight is to reduce the problem of synthesizing QRE lenses to the problem of
synthesizing {\em bijective lenses}~\cite{optician}, a simpler subset of
lenses that support strict bijections, rather than bijections modulo
equivalence relations.

@@

This technical investigation justifies replacing manually constructed lenses that may use
functional composition and nested canonizers with synthesized lenses in normal form.
The end result is that users may synthesize complex transformations with a single line
of code.  For instance, all the code in Figure~\ref{fig:example-lens} may be replaced by
a single call to the synthesis prodedure:

\begin{lstlisting}
let bib_to_end : (lens in bibtex $\Leftrightarrow$ endnote)
= synth bibtex $\Leftrightarrow$ endnote  using  {(bib_example, end_example)}
\end{lstlisting}
\noindent Here, the generated  quotient lens synchronizes \cd{bibtex} and
\cd{endnote} formats.  It uses \cd{bib_example} and \cd{end_example} (the two concrete example
strings given at the beginning of this section) to disambiguate.
In additiona, and as we saw earlier with the definition of \cd{s_to_c}, the synthesis procedure itself
can be used to create lenses that are in turn used to define other QREs.  The ability to intermix
QRE specification with QRE lens synthesis yields a powerful and flexible way of
creating bidirectional transformations.

\section{Quotient Regular Expressions}
\label{QRE}

Quotient Regular Expressions allow programs to specify a regular language,
an equivalence relation over elements of the language and a function to convert
elements of the language into normal form.

\subsection{Syntax}

The language of Quotient Regular Expressions (QREs) is given by the following
grammar:
\begin{align*}
c := \; &R \sep \project{R}{s} \sep \squash{c}{c'}{f} \sep
\perm{c_1, \ldots, c_n}{c} \\
& | \; \normalize{R}{R'}{f} \sep c \semicolon c' \sep c \cdot c''' \sep (c \sep
c') \sep c^*,
\end{align*}
where $c$ ranges over QREs, $R$ ranges over regular expressions, $f$ ranges over
functions between regular languages, and $s$ ranges over character strings.
$\mathcal{L}(R)$ is
the language accepted by the regular expression $R$.

%Section~\ref{subsec:qre-expressions} presented a number of examples

\subsection{Overview of Quotient Regular Expression}
In the \bibtex{} to EndNote example, we encountered the $\project{R}{s}$,

\noindent $\squash{c}{c'}{f}$ and $\perm{c_1, \ldots, c_n}{c}$ QREs, as well as
the regular operators applied to these, as in

\begin{lstlisting}
let space_name  = name | (name . wsp_to_space . (name . wsp_to_space)$^*$ . name)
\end{lstlisting}

One combinator that we haven't yet introduced is sequential composition.
The $c \semicolon c'$ QRE successively applies
two equivalence relations to a regular language. For example, suppose that
we wish to express a name as $\kw{last\_name \cdot ``,\text{''} \cdot
\text{wsp+} \cdot first\_name}$, or as $\kw{first\_name\cdot \text{wsp+} \cdot
last\_name}$, with $\kw{first\_name\cdot \text{wsp+} \cdot last\_name}$ as the
canonical representation. We also wish to project away the extra whitespace in
between \kw{first\_name} and \kw{last\_name} once this identification has been
made. Then we can express this equivalence relation by a composition of a
\textit{squash} QRE and a QRE that projects away the extra whitespace in between
\kw{first\_name} and \kw{last\_name}:

\begin{lstlisting}
(* Regular expressions matching different comma and space representations *)
let comma_name = last_name . "," . wsp . first_name
let space_name = first_name . wsp . last_name

(* Squash comma representation into space representation *)
let comma_to_space =
synth comma_name <=> space_name using {("Conway, J. H.", "J. H. Conway")}
let comma_to_space = squash comma_name $\to$ space_name using (get comma_to_space)

(* Project away whitespace in space_name representation *)
let project_in_space = first_name . (project wsp+ $\to$ " ") . last_name

(* Compose the two QREs *)
let valid_name = project_in_space ; comma_to_space
\end{lstlisting}

The final combinator is $\normalize{R}{R'}{f}$, which expresses
the equivalence relation that sends each element in $\mathcal{L}(R)$ to its
canonical representation $f(r) \in \mathcal{L}(R')$. The equivalence relation
defined by $\normalize{R}{R'}{f}$ is thus the equivalence relation on
$\mathcal{L}(R)$ defined by the {\em fibres} of $f$ (i.e. $r \sim r'$ if and
only if $f(r) = f(r'))$. The canonizing function $f$ is required to be
surjective and idempotent.

The $\normalize{R}{R'}{f}$ QRE is the most general QRE in that for each QRE
$q$, there exist regular expressions $R, R'$ and a surjective, idempotent
function $f:\mathcal{L}(R) \longrightarrow \mathcal{L}(R')$ such that $q$ and
$\normalize{R}{R'}{f}$ each define the same equivalence relation on
$\mathcal{L}(R)$, with $\mathcal{L}(R')$ forming a complete set of
representatives for this equivalence relation. The $\normalize{R}{R'}{f}$
combinator is included so as to express equivalence relations that are either
difficult to define, or that cannot be defined using the other combinators.
\saz{Add something about being ``complete'' here?}\sam{I'm not sure how to do
this here. The precise statement about how normalize is strong enough to
express all QREs requires that we introduce W(c), K(c) and Canonizer(c) which
we haven't introduced at this point in the paper}

Now in order for the programmer to ensure that the {\em normalize} QRE is well
formed, she must define a canonizing function $f$ between two regular
languages and prove that $f$ is surjective and idempotent: the compiler cannot
verify this since this problem is undecidable in general. Consequently, we do
not use the $\normalize{R}{R'}{f}$ QRE in practice. On the other hand, the
$\normalize{R}{R'}{f}$ combinator gives a sufficient condition for potential
QREs to be ``well-behaved'', particularly with respect to our strategy for
synthesizing quotient lenses which we shall introduce in Section~\ref{synth}.

The equivalence relations definable by QREs are a strict subset of those that
can be expressed in Boomerang. For instance, in Boomerang, one can use any lens
to canonize a regular language described by $R$ to a regular language described
by $S$, with $\mathcal{L}(S) \not \subseteq \mathcal{L}(R)$, whereas each QRE
$c$ canonizes a regular language described by $R$ to regular subset
$\mathcal{L}(R') \subseteq \mathcal{L}(R)$ described by $R'$. QREs therefore are
not as expressive as Boomerang canonizers. On the other hand, QREs enable us to
express many equivalences that occur in real world data as we shall demonstrate
in the evaluation section (Section~\ref{impl}).

The net effect of these design decisions means that each QRE $c$ enables
us to simultaneously express
\begin{enumerate}
  \item a regular expression $W(c)$ (the ``whole'' of $c$),
  \item an equivalence relation $\eqrel{c}$ on $\mathcal{L}(W(c))$,
  \item a regular expression $K(c)$ (the ``kernel'' of $c$)
  such that $\mathcal{L}(K(c))$ forms a complete set of representatives for
  $\eqrel{c}$, and
  \item a ``canonizing'' function $\canonizer(c):\mathcal{L}(W(c))
  \longrightarrow \mathcal{L}(K(c))$ which given any $w \in \mathcal{L}(W(c))$,
  computes $\canonizer(c)(w)$ as the unique $k$ in $\mathcal{L}(K(c))$ such that
  $k$ is equivalent to $w$ mod $\eqrel{c}$
\end{enumerate}
Observe that since $\canonizer(c)$ computes $\canonizer(c)(w)$ as the unique
$k$ in $\mathcal{L}(K(c))$ such that $w$ is equivalent to $k$ mod $\eqrel{c}$,
then the equivalence classes of $\eqrel{c}$ are the same as the fibres of
$\canonizer(c)$.

>>>>

To begin, consider a bidirectional transformation that converts
\bibtex{} citation records such as
\begin{verbatim}
@Book {Doe,
Author = "John Doe",
Title = {Generic Title}
}
\end{verbatim}
\noindent
into an equivalent EndNote format like the following
\begin{verbatim}
%0 Book
%T Generic Title
%A John Doe
%F Doe
\end{verbatim}
\noindent
and vice versa.
%
Notice that transformations from one format to the other preserve
most of the information present in the data:  the title, the author,
and the label, for instance.  However, some of the syntactic information
may be lost.  For instance, users are typically uninterested in preserving
whitespace between words or
between fields.  The order of author and title fields is also likely
irrelevant and there may be equivalent ways of writing the same
name:  ``Doe, John'' vs ``John Doe.''

%% Moreover,
%% if two \bibtex{} files differ only in ``inessential'' details, we would like
%% to treat them the same by mapping them to identical EndNote representations.
%% In this , the transformation should allow whitespace characters to occur
%% arbitrarily between the words in the label, author, and title fields, as well as
%% in between them. Second, an author's name can have the surname followed by a
%% comma and then the other names separated by arbitrary whitespace, or just all
%% the names separated by arbitrary whitespace (for example, the names ``Doe, J.
%% H.'' and ``John Doe'' are equivalent). Lastly, the transformation should
%% also ignore the order in which the author and title fields occur in each
%% citation record.

When two instances of the same format differ only terms of such inessential
information, it is often useful to treat them as identical and to map them
into canonical representations.
Continuing our example, the following two
\bibtex{} citations represent the same logical object and as a result, should
be translated to the same EndNote record.
\begin{verbatim}
@Book {Doe,
Author = "John Doe",
Title = {Generic Title},
}

@Book{     Doe,
Title = {Generic Title},
Author = "Doe, John",     }
\end{verbatim}

@@

@@

\begin{lstlisting}
let p_wsp = project wsp+ $\to$ " "
\end{lstlisting}

Sometimes there are multiple disjoint representations of the same data.
In such situations, the \kw{squash} combinator creates a QRE that
allows external data to be in either format, and converts any
data in one format to the other.
For instance, assume
the \codefont{comma\_name} format describes ``Doe, John''
and the \codefont{space\_name} format describes ``John Doe''
and \codefont{s\_to\_c} is a function from the first to the second.  In this case,
the following instance of \kw{squash} creates the desired canonizer.

@@

One way to define the function used by squash is simply to write it
from scratch in OCaml.  However, our system will also synthesize
such functions automatically from pairs of regular languages and
examples.
%
\begin{lstlisting}
let l = synth comma_name $\Leftrightarrow$ space_name using {("Doe, John", "John Doe")}
let s_to_c = get l
\end{lstlisting}
%
The first line above synthesizes a \emph{bidirectional} lens between
\codefont{comma\_name} and \codefont{space\_name} using the listed example
transformation as a guide.  The second line simply
extracts the get direction transformation from the lens,
which is what we need for squash.

Another QRE primitive is the composition combinator ``;''. This combinator
applies two equivalence relations to a regular language in sequence. For
example, suppose that we wish to express a name as $\kw{last\_name \cdot
``,\text{''} \cdot \text{wsp+} \cdot first\_name}$, or as $\kw{first\_name\cdot
\text{wsp+} \cdot last\_name}$, with $\kw{first\_name\cdot \text{wsp+} \cdot
last\_name}$ as the canonical representation. We also wish to project away the
extra whitespace in between \kw{first\_name} and \kw{last\_name} once this
identification has been made. Then we can express this compound equivalence
relation by a composition of a \textit{squash} QRE and a QRE that projects away
the extra whitespace in between \kw{first\_name} and \kw{last\_name}

A third useful QRE primitive is the permutation combinator.  For example,
the following instance of \kw{perm} allows label, author and title fields
(which we assume have been defined earlier) to appear in any order.
%
\begin{lstlisting}
let bib_fields = perm (label, bib_author, bib_title)
\end{lstlisting}
%
To normalize any repeated separator between fields, one can extend a permutation
with any other QRE.  For instance, below, we normalize whitespace between fields,
leaving only a single newline.
%
\begin{lstlisting}
let bib_fields = perm (label, bib_author, bib_title) with (project (wsp+ . "," . wsp+) $\to$ ",\n")
\end{lstlisting}
%
The final QRE combinator is the normalize combinator.
This combinator allows a programmer to manually define a function
$f$ which sends each string that matches a regular expression $R$ to some
canonical representative in another regular expression $R'$. The equivalence
relation defined by the normalize combinator is hence the equivalence relation
defined by the {\em fibres} of $f$; that is, for all strings $s$ and $s'$ that
match $R$, $s$ is equivalent to $s'$ if and only if $f(s) = f(s')$.

For instance, assume that $f(s)=$ `` '' for all whitespace strings $s$. Then the
project QRE \kw{p_wsp} which matches any non-zero number of whitespace
characters and converts any such whitespace into a single space character can
be expressed using the normalize combinator as
\begin{lstlisting}
let p_wsp = normalize (wsp+, " ", f)
\end{lstlisting}
The QRE for the bibtex format is presented in its entirety in
Figure~\ref{fig:example-qre}.
\begin{figure}[t]
\begin{lstlisting}
let wsp = [ \n\t\r]

(* define name representations with a space and with a comma *)
let space_name = first_name . wsp+ . last_name
let comma_name = last_name . "," . wsp+ . first_name

(* synthesize a lens that maps comma representation to space representation *)
let l = synth comma_name $\Leftrightarrow$ space_name using {("Doe, John", "John Doe")}
let s_to_c = get l

(* squash QRE maps comma_name to space_name *)
let comma_to_space = squash comma_name $\to$ space_name using (get comma_to_space)

(* map arbitrary whitespace into " " as a canonical representative *)
let p_wsp = project wsp+ $\to$ " "

(* Project away whitespace in space_name representation *)
let project_in_space = first_name . p_wsp . last_name

(* Compose the two QREs to get a canonical representation for a name *)
let name = project_in_space ; comma_to_space

(* define rest of bibtex fields *)
let name_sep = p_wsp . "and" . p_wsp
let bib_names = name | (name . name_sep . (name . name_sep)$^*$ . name)
let bib_author = "author = \"" . bib_names . "\""
let title = word | (word . p_wsp . (word . p_wsp)$^*$ . word)
let bib_title = "title = {" . title . "}"

(* allow any permutation of fields interspersed with arbitrary whitespace *)
let bib_fields = perm (label, bib_author, bib_title) with (project (wsp+ . "," . wsp+) $\to$ ",\n")
let bibtex = "@book{" . bib_fields . "}"
\end{lstlisting}
\caption{QRE specification of \bibtex{} records. }
\label{fig:example-qre}
\end{figure}
\subsection{QRE Lenses}
QREs allow users to specify regular languages modulo equivalence relations,
a first step in writing many data transformations.  However, writing the
inner bijective lens itself still takes a fair amount of work.  Each such lens
is constructed from simple primitive combinators including:
%
\begin{itemize}
  \item \kw{del} $s$: delete the constant string $s$ in the get direction;
  produce the constant string $s$ in the put direction.
  \item \kw{ins} $s$: produce the string $s$ in the get direction;
  delete the string $s$ in the put direction.
  \item \kw{copy} $R$: copy the text matching the regular expression $R$ in both
  directions.
\end{itemize}
In addition, such primitives may be combined using the regular operators
Kleene star, alternation and concatenation, or functional composition of
lenses.  Figure~\ref{fig:example-lens} illustrates the use of these
combinators to transform \bibtex{} to EndNote (and vice versa).
A fair amount of trickiness in writing these transformations as users
must keep mental track of what they have parsed so far \emph{in both
directions}, carefully inserting and deleting the relevant syntax on both
sides.

@@

let author_lens : (lens in (",\n" . bib_author) $\Leftrightarrow$ ("\n" . end_author)) =
    del ",\nauthor = \""
        . ins "\n%A "
        . name
        . (del " and "
        . ins "\n"
        . (ins "%A " . name . del " and " . ins "\n")* 
        . ins "%A "
        . name
        || "")

@@

let bib_to_end : (lens in bibtex $\Leftrightarrow$ end_note) =
    preamble . label . author_lens . title_lens
\end{lstlisting}
\caption{An example lens between \cd{bibtex} and \cd{end_note} regular
expressions with no equivalences.}
\label{fig:example-lens}
\end{figure}

%% The lens shown in Figure~\ref{fig:example-lens} transforms \bibtex{} data to
%% EndNote data  (and vice versa) by matching the delimiters and separators in a
%% bijective manner. For example, the \cd{preamble} lens replaces the \bibtex{}
%% preamble \verb|@Book| with the EndNote preamble \verb|%0 Book\n%F|. When the
%% programmer is writing the lens, she has to make sure that all of these small
%% deletions and insertions are correct, which can be a challenging task if the
%% lens is large enough, as in the definition of the lens \verb!author_lens! in
%% Figure~\ref{fig:example-lens}. This makes lens programming tedious and
%% error-prone.
Consequently, we offer the programmer some relief by identifying a class of
quotient lenses that we call {\em QRE lenses}, and by showing how to {\em synthesize}
QRE lenses from a pair of QREs and a set of input-output example pairs. Our key
insight is to reduce the problem of synthesizing QRE lenses to the problem of
synthesizing {\em bijective lenses}~\cite{optician}, a simpler subset of
lenses that support strict bijections, rather than bijections modulo
equivalence relations.

@@

This technical investigation justifies replacing manually constructed lenses that may use
functional composition and nested canonizers with synthesized lenses in normal form.
The end result is that users may synthesize complex transformations with a single line
of code.  For instance, all the code in Figure~\ref{fig:example-lens} may be replaced by
a single call to the synthesis prodedure:
\begin{lstlisting}
let bib_to_end : (lens in bibtex $\Leftrightarrow$ endnote)
= synth bibtex $\Leftrightarrow$ endnote  using  {(bib_example, end_example)}
\end{lstlisting}
\noindent Here, the generated  quotient lens synchronizes \cd{bibtex} and
\cd{endnote} formats.  It uses \cd{bib_example} and \cd{end_example} (the two
concrete example strings givn at the beginning of this section) to disambiguate.
In addition. as we saw earlier with the definition of \cd{s_to_c}, the synthesis
procedure itself can be used to create lenses that are in turn used to define
other QREs. The ability to mix QRE specification with QRE lens synthesis
yields a powerful and flexible way of creating bidirectional transformations.

\section{Quotient Regular Expressions}
\label{QRE}
\subsection{Syntax and Semantics of QREs}
Formally, the language of Quotient Regular Expressions (QREs) is given by the
following grammar:
\begin{align*}
c := \; &R \sep \project{R}{s} \sep \squash{c}{c'}{f} \sep
\perm{c_1, \ldots, c_n}{c} \\
& | \; \normalize{R}{R'}{f} \sep c \semicolon c' \sep c \cdot c''' \sep (c \sep
c') \sep c^*,
\end{align*}
where $c$ ranges over QREs, $R$ ranges over regular expressions, $f$ ranges over
functions between regular languages, and $s$ ranges over character strings.

Using the conventional notation that $\mathcal{L}(R)$ is the language accepted
by the regular expression $R$, then each QRE $c$ enables us to express
\begin{enumerate}
  \item a regular expression $W(c)$ (the ``whole'' of $c$),
  \item an equivalence relation $\eqrel{c}$ on $\mathcal{L}(W(c))$,
  \item a regular expression $K(c)$ (the ``kernel'' of $c$)
  such that $\mathcal{L}(K(c))$ forms a complete set of representatives for
  $\eqrel{c}$, and
  \item a ``canonizing'' function $\canonizer(c):\mathcal{L}(W(c))
  \longrightarrow \mathcal{L}(K(c))$ which given any $w \in \mathcal{L}(W(c))$,
  computes $\canonizer(c)(w)$ as the unique $k$ in $\mathcal{L}(K(c))$ such that
  $k$ is equivalent to $w$ mod $\eqrel{c}$
\end{enumerate}
Figures~\ref{fig:wk} gives the inductive definitions for the whole and
kernel languages $W(c)$ and $K(c)$ of a QRE.
\begin{figure}[t]
\centering
\[
\begin{array}{l@{\quad}l@{\quad}l}

c & W(c) & K(c) \\ \hline
R & R & R \\
\project{R}{s} & R & s \\
\squash{c}{c'}{f} & W(c) \sep W(c') & K(c') \\
\normalize{R}{R'}{f} & R & R' \\
c_2 \; ; \; c_1 & W(c_1) & K(c_2) \\
c_1 \cdot c_2 & W(c_1) \cdot W(c_2) & K(c_1) \cdot K(c_2) \\
c_1 \sep c_2 & W(c_1) \sep W(c_2) & K(c_1) \sep K(c_2) \\
c^* & W(c)^* & K(c)^* \\
\end{array}
\]
\[
\begin{array}{r@{\quad}l}
W( \perm{c_1, \ldots, c_n}{c} ) = &
\bigcup \limits_{\sigma \in S_n} W(c_{\sigma(1)}) \cdot W(c) \cdot \ldots \cdot
W(c) \cdot W(c_{\sigma(n)})\\
K( \perm{c_1, \ldots, c_n}{c} ) = & K(c_1) \cdot K(c) \cdot \ldots \cdot K(c)
\cdot K(c_n)
\end{array}
\]
\caption{Whole and Kernel Regular Expressions}
\label{fig:wk}
\end{figure}
The \textit{squash} and \textit{permutation} combinators give the two
interesting definitions. If $c = \squash{c}{c'}{f}$, then the whole language of
$c$ is $W(c) \sep W(c')$. This is because the \textit{squash} combinator takes
the whole language $W(c)$ of $c$, the whole language $W(c')$ of $c'$ and a
function $f : \mathcal{L}(W(c)) \longrightarrow \mathcal{L}(W(c'))$, maps
$\mathcal{L}(W(c))$ into $\mathcal{L}(W(c'))$ using $f$ and then canonizes
$W(c')$ into $K(c')$ using $\canonizer(c')$.

For the $\perm{c_1, \ldots, c_n}{c}$ combinator, the whole language is the
union of languages of the form $W(c_{\sigma(1)}) \cdot W(c) \cdot \ldots \cdot
W(c) \cdot W(c_{\sigma(n)})$. This is because the \textit{permutation}
combinator allows for the string to match any permutation the $c_i$'s while also
taking into account the separator $c$ in between each of the $c_i$'s. The kernel
of the \textit{permutation} combinator is the language $K(c_1) \cdot K(c) \cdot
\ldots \cdot K(c) \cdot K(c_n)$, because the canonical permutation is the
identity permuation, with each of the parts of the string that match $c_i$
and $c$ canonized into $K(c_i)$ and $K(c)$ respectively.

Figure~\ref{fig:canonizers} gives the inductive definitions of the
$\canonizer{}$ function for each QRE.
\begin{figure}[t]
\begin{center}
\[
\begin{array}{l@\quad l @\quad l}
\canonizer(R) &=& id_{\mathcal{L}(R)} \\
\canonizer(\project{R}{s})(w) &=& s\\
\canonizer(\squash{c}{c'}{f})(w) &=&
\begin{cases}
\canonizer(c)(f(w)) & \text{if } w \in \mathcal{L}(W(c))\\
\canonizer(c')(w) & \text{otherwise}
\end{cases}\\
\canonizer(\normalize{R}{R'}{f}) &=& f\\
\canonizer(c' \; ; \; c) &=& \canonizer(c') \circ \canonizer(c)\\
\canonizer(c' \cdot c) &=& \canonizer(c') \cdot \canonizer(c)\\
\canonizer(c_1 \sep c_2)(w) &=&
\begin{cases}
\canonizer(c_1)(w) & \text{if } w \in \mathcal{L}(W(c_1))\\
\canonizer(c_2)(w) & \text{if } w \in \mathcal{L}(W(c_2))\\
\end{cases}\\
\canonizer(c^*) &=& \canonizer(c)^* \\
\end{array}
\]
\end{center}
\begin{align*}
&\canonizer(\perm{c_1, \ldots, c_n}{c})(r_{\sigma(1)}
\cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot r_{\sigma(n)})\\
&= (\canonizer(c_1)(r_1)) \cdot (\canonizer(c)(s_1)) \cdot \ldots \cdot
(\canonizer(c_n)(r_n))
\end{align*}
\caption{QRE canonizers}
\label{fig:canonizers}
\end{figure}
Again, the \textit{permutation} combinator gives the most interesting
definition,
\begin{align*}\canonizer(\perm{c_1, \ldots, c_n}{c})(r_{\sigma(1)}
\cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot r_{\sigma(n)}) \\
= (\canonizer(c_1)(r_1)) \cdot (\canonizer(c)(s_1)) \cdot \ldots \cdot
(\canonizer(c_n)(r_n))
\end{align*}
\noindent which as we have explained is defined so that the strings that match
$c_i, c$ are placed in the canonical permutation which is the identity
permutation before $\canonizer(c_i)$ and $\canonizer(c)$ are applied
respectively.
